---
title: "Tidy Data"
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

Rules for Tidy Data: 

* Columns are variables 
* Rows are observations 
* Every value has a cell

Relational Data:

* Data spread across tables with defined relations 
* Variables used to define these relations are KEYS 
* Tables are combined by JOINS

```{r setup, message=FALSE}
library(tidyverse)
```

### Function `pivot_longer`

Load the PULSE data

```{r}
pulse_data =
    haven::read_sas("C:/Users/Twilight/Desktop/Columbia/Fall Semester 2024/Data Science I/Datasets/public_pulse_data.sas7bdat") %>%
    janitor::clean_names()

# Standardizes Column Names:
#   Converts all column names to lowercase.
#   Replaces spaces, special characters, and other problematic symbols with underscores (_).
#   Removes accents from letters (if any).
#   Ensures that column names begin with a letter and are syntactically valid R names.
# Results in Consistent and Tidy Names:
#   Helps avoid issues with inconsistent column names, which can be particularly useful when working with data from external sources like Excel files or databases.
#   Works well for making column names easier to work with programmatically (e.g., when selecting columns or using functions in dplyr or ggplot2).

```

### Wide format to long format ...

```{r}
pulse_data_tidy = pulse_data %>%
    pivot_longer(
      bdi_score_bl:bdi_score_12m,
      names_to = "visit",
        names_prefix = "bdi_score_",
      values_to = "bdi"
    )

# Prepare to convert conlumns bdi_score_bl:bdi_score_12m
# Put names into newly created "visit" column, and put values into newly created "bdi" column
  # In column "visit", only keep original column names after the names_prefix "bdi_score_"

```

### Rewrite, combine, and extend(to add a mutate)

```{r}
pulse_data = 
    haven::read_sas("C:/Users/Twilight/Desktop/Columbia/Fall Semester 2024/Data Science I/Datasets/public_pulse_data.sas7bdat") %>%
    janitor::clean_names() %>%
    pivot_longer(
        bdi_score_bl:bdi_score_12m,
        names_to = "visit",
          names_prefix = "bdi_score_",
        values_to = "bdi"
    ) %>%
    relocate(id, visit) %>%  # id与visit互换位置
    mutate(visit = recode(visit, "bl"="00m"))  # visit中的"bl"换成"00m"
```

### Function `pivot_wider`

```{r}
analysis_result = 
    tibble(
      group = c("treatment", "treatment", "placebo", "placebo"),
      time  = c("pre", "post", "pre", "post"),
      mean  = c(4, 8, 3.5, 4)
    )

analysis_result2 = 
  analysis_result %>%
    pivot_wider(
        names_from  = "time",
        values_from = "mean"
    )
        
analysis_result
analysis_result2

# extract names from "time" to become the new columns
# extract values from "mean" to become the new cell contents

```

### Binding rows

First step: import each table.
```{r}
fellowship_ring = 
  readxl::read_excel("C:/Users/Twilight/Desktop/Columbia/Fall Semester 2024/Data Science I/Datasets/LotR_Words.xlsx", range = "B3:D6") |>
  mutate(movie = "fellowship_ring")

two_towers = 
  readxl::read_excel("C:/Users/Twilight/Desktop/Columbia/Fall Semester 2024/Data Science I/Datasets/LotR_Words.xlsx", range = "F3:H6") |>
  mutate(movie = "two_towers")

return_king = 
  readxl::read_excel("C:/Users/Twilight/Desktop/Columbia/Fall Semester 2024/Data Science I/Datasets/LotR_Words.xlsx", range = "J3:L6") |>
  mutate(movie = "return_king")

```

Bind all the rows together
```{r}
lotr_tidy = 
    bind_rows(fellowship_ring, two_towers, return_king) %>%
    janitor::clean_names() %>%
    relocate(movie) %>%  # 把movie列放在第一列
    pivot_longer(
        female:male,
        names_to = "gender",
        values_to = "words"
    )
```

### Joining Tables

Import and Clean the FAS datasets.
```{r}
pups_df =
    read.csv("C:/Users/Twilight/Desktop/Columbia/Fall Semester 2024/Data Science I/Datasets/FAS_pups.csv") %>%
    janitor::clean_names() %>%
    mutate(sex=recode(sex, `1`="male", `2`="female"))

litters_df =
    read_csv("C:/Users/Twilight/Desktop/Columbia/Fall Semester 2024/Data Science I/Datasets/FAS_litters.csv") %>%
    janitor::clean_names() %>%
    relocate(litter_number) %>%
    separate(group, into=c("dose", "day_of_tx"), sep=3) 
  #原本group列内容为Con7,现在在第三字节处切割，第一部分(Con)给到"dose"列，第二部分(7)给到"day_of_tx"列
```

Next up, time to join them!
```{r}
fas_df =
    left_join(pups_df, litters_df, by="litter_number") %>%
    arrange(litter_number) %>%   # 将litter_number列升序排列，若有N/A项优先
    relocate(litter_number, dose, day_of_tx) # 将litter_number, dose, day_of_tx放在1，2，3列
```

